---
journal: "Geophysical Research Letters"
options: "draft,linenumbers"
title: "How stationary are planetary waves in the Southern Hemisphere?"
authors:
  - name: Elio Campitelli
    affil: 1
  - name: Carolina Vera
    affil: "1, 2"
  - name: Leandro Díaz
    affil: "1, 2"
affiliations:
  - number: 1
    name: "Centro de Investigaciones del Mar y la Atmosfera, UMI-IFAECI (CONICET-UBA-CNRS)"
  - number: 2
    name: "Departamento de Ciencias de la Atmósfera y los Océanos (FCEyN, UBA)"
corresponding_author:
  - name: Elio Campitelli
    email: elio.campitelli@cima.fcen.uba.ar
keypoints:  
  - "Zonal waves and Quasi-stationary waves are disctinct but related phenomena"
  - "This distinction has theoretical and practical implications"
  - "The relationship between the mean ZW amplitude and QS amplitude yields an estimate of stationarity"
abstract: >
   Abstract goes here
knit: (function(file, encoding, ...) {
    rmarkdown::render(file, encoding = encoding, output_format = "all")})
output:
   rticles::agu_article:
      keep_tex: true
   word_document:
      default
bibliography: [qszw.bib,qszw-pkgs.bib]
header-includes: 
      - \usepackage{gensymb} 
      - \usepackage{soulutf8}
      - \usepackage{subfig}
params: 
  draft: TRUE
---


```{r setup, include=FALSE}
# Some recommended settings. 
knitr::opts_chunk$set(
   echo = FALSE,
   fig.pos = 'h',
   out.extra = "",   # To force the use of figure enviroment
   fig.cap = "Please caption every figure"
)

# Notification
start.time <- unclass(Sys.time())
min.time <- 10
knit_doc <- knitr::knit_hooks$get("document")
knitr::knit_hooks$set(document = function(x) {
   took <- unclass(Sys.time()) - start.time
   if (unclass(Sys.time()) - start.time >= min.time) {
      notify("Done knitting!", 
             paste0("Took ", round(took), " seconds"),
             time = 5)
   }  
   knit_doc(x)
})

name <- tools::file_path_sans_ext(knitr::current_input())
knitr::opts_chunk$set(cache = TRUE,
                      warning = FALSE, 
                      message = FALSE,
                      cache.extra = 2,
                      cache.path = paste0("cache/", name, "/"),
                      fig.path = paste0("fig/", name, "/"),
                      fig.align = "center")

library(metR)
library(data.table)
library(ggplot2)
library(metR)
library(magrittr)
library(circular)
library(RcppRoll)
library(patchwork)
library(lubridate)
library(ggperiodic)
library(latex2exp)
library(glue)

here <- here::here

gluec <- function(...) {
   unname(vapply(c(...), function(x) glue(x), FUN.VALUE = "a"))
}
source(here("scripts/helperfun.R"))
remove(ggplot)

data.world <- BuildMap(res = 1, smooth = 1)
map.world <- geom_map2(data.world)
map.SH <- geom_map2(data.world[lat %b% c(-90, 20)], color = "gray20")

lev.breaks <- c(1000, 500, 300, 200, 100, 50, 10)

theme_elio <- theme_minimal(base_size = 9) +
   theme(
      # text = element_text(family = font_rc),
      legend.position = "bottom", legend.box = "vertical",
      panel.spacing.y = unit(5, "mm"),
      panel.spacing.x = unit(5, "mm"),
      legend.spacing = unit(2, "mm"),
      plot.margin = grid::unit(rep(3, 4), "mm"),
      legend.title = element_blank(),
      legend.box.spacing = unit(3, "mm"),
      legend.margin = margin(t = -5),
      panel.grid = element_line(color = "gray50", size = 0.2, linetype = 3),
      panel.ontop = TRUE)
theme_set(theme_elio)
guide_colorstrip_bottom <- function(width = 25, height = 0.5, ...) {
   guide_colorstrip(title.position = "top", title.hjust = 0.5,
                    barheight = height,
                    barwidth = width, ...)
}
# 
# knitr::write_bib(c("base", "ggplot2", "metR", "data.table"), 
#                  "qszw-pkgs.bib")

session_file <- "session_info.txt"
file.create(session_file) 
op <- options(width = 100)
writeLines(capture.output(sessioninfo::session_info()), session_file)
options(op)
```

# Introduction

Zonal asymmetries of extratropical circulation in the Southern Hemisphere (hereafter called as “planetary waves”) strongly modulate weather systems and regional climate through latitudinal transport of heat, humidity, and momentum (REFS), and by contributing to the development of blocking events [e.g. @Trenberth1985]. 

In Rossby wave theory, stationary waves are those with zero frequency or phase velocity [@Holton2012a]. In practice, however, most studies have assumed the “quasi-stationary” nature of southern hemisphere planetary waves based on @Loon1972. In this foundational study, the authors analyzed data from 1957 and 1958 and found that while extratropical waves with wavenumber 1 to 6 had comparable amplitudes in daily fields, only wavenumbers 1 and 3 contributed significantly to the climatological field. From that, they concluded that only waves 1 and 3 recur consistently in the same location and thus have a significant quasi-stationary component on top of a "moving" component. This was a qualitative conclusion and to our knowledge no study has quantified the level of stationarity of each wavenumber. 

After more than four decades from the publication of @Loon1972, and considering the current availability of different global reanalysis datasets, in this study we assess the stationarity features of planetary waves in the southern hemisphere, and extends @Loon1972's methodology into a quantitative measure of planetary wave stationarity, which we apply to both hemispheres. 

# Methods

## Planetary waves

We define *planetary waves* as waves that extend along a full latitude circle. *Zonal waves* (ZW) are planetary waves of the “instantaneous” fields and  *quasi-stationary waves* (QS), planetary waves of the time-mean field such that: 

\begin{linenomath*}
\begin{eqnarray}\label{eq:ZW}
\mathrm{ZWk}(t) & = & A_\mathrm{ZWk}(t)\cos \left [ k\lambda - \alpha_\mathrm{ZWk}(t) \right ] \\ 
\overline{\mathrm{ZWk}(t)} = \mathrm{QSk} & = & A_\mathrm{QSk}\cos \left (  \mathrm{k}\lambda - \alpha_\mathrm{QSk} \right ) \label{eq:QS}
\end{eqnarray}
\end{linenomath*}

where $\mathrm{k}$ is wavenumber, $\lambda$ longitude, and $\mathrm{A_{x}}$ and $\alpha_\mathrm{x}$, amplitude and phase, respectively. $\mathrm{ZWk}(t)$ depends on time, but not $\mathrm{QSk}$. 

These definitions depend on which are the "instantaneous fields" and the averaging time-scales. A dataset of 365 daily mean fields defines 365 daily zonal waves and one annual quasi-stationary wave but 12 monthly quasi-stationary waves (per level and latitude). A 30 year dataset of monthly mean fields define 360 monthly zonal waves and one 30-year quasi-stationary wave. Monthly planetary waves are quasi-stationary waves in one case and zonal waves in the other. 

```{r read-rao}
period_rao <- c(1948, 2017)
N <- diff(period_rao) + 1
lats <- 50
period_cap_rao <- glue("From monthly NCEP/NCAR Reanalysis, {period_rao[1]} to {period_rao[2]}")
rao_data <- rbind(ReadNetCDF(here("DATA/NCEP Reanalysis/hgt.mon.mean.nc"), 
                             c(gh = "hgt") ,
                             subset = list(lat = -lats, 
                                           time = gluec("{period_rao[1]}-01-01",
                                                        "{period_rao[2]}-12-31"))),
                  ReadNetCDF(here("DATA/NCEP Reanalysis/hgt.mon.mean.nc"), 
                             c(gh = "hgt") ,
                             subset = list(lat = lats, 
                                           time = gluec("{period_rao[1]}-01-01",
                                                        "{period_rao[2]}-12-31"))))
setnames(rao_data, "level", "lev")
```

## Stationarity 

From the properties of the superposition of waves we can deduce that, in general, the stationary phase $\alpha_\mathrm{QSk}$ does not equal $\overline{\alpha_\mathrm{ZWk}}$ and the stationary amplitude $A_\mathrm{QSk}$ is less or equal $\overline{A_\mathrm{ZWk}}$ [@Pain2005]. 

We use this latter property and use the quotient between $A_\mathrm{QS}$ and $\overline{A_\mathrm{ZW}}$ as a measure of quasi-stationary wave stationarity such as:

\begin{linenomath*}
\begin{equation}\label{eq:S}
\hat{S} = \frac{A_\mathrm{QS}}{\overline{A_\mathrm{ZW}}}
\end{equation}
\end{linenomath*}

For a sample of $n$ completely random waves, the expected value of $\hat{S}$ is $n^{-1/2}$ because the average amplitude of the sum of $n$ waves with random phases and mean amplitude $A$ is $An^{-1/2}$ [@Pain2005]. For completely stationary waves $\hat{S} = 1$ irrespective of sample size. 

$\hat{S}$ is used --sometimes as $2/\pi\arcsin \left (\hat{S} \right )$ [@Singer1967]-- in the meteorological literature in the context of wind steadiness [e.g @Hiscox2010]. To our knowledge this is the first time it has been applied to the study of atmospheric waves. 

Equation \ref{eq:S} is equivalent to

\begin{linenomath*}
\begin{equation}\label{eq:S2}
\hat{S} =   \frac{\sum_t A_\mathrm{ZW}(t) \cos  \left [\alpha_\mathrm{zw}(t) - \alpha_{qs} \right ]}{\sum_t A_\mathrm{ZW}(t)}
\end{equation}
\end{linenomath*}

The numerator represents the sum of the zonal waves amplitudes projected onto the direction of the quasi-stationary wave. Waves that deviate from that direction decrease the overall stationarity in proportion to their amplitude. This definition of stationarity depends on the phase distribution and its relationship with amplitude instead of the phase velocity of waves. Thus, it's a statistical --rather than dynamical-- property. 

We used Equation \ref{eq:S2} to compute a timeseries of quasi-stationary wave stationarity. We first calculated $\alpha_{qs}$ for each month and then applied Equation \ref{eq:S2} with a 15-year rolling window approximated using loess regression with degree 0.

## Data

We use monthly geopotential fields from the NCEP/NCAR Reanalysis [@Kalnay1996] for the period `r period_rao[1]` to `r period_rao[2]` and compute one quasi-stationary wave for the whole period for each month, level and wavenumber. Amplitude and phase for each wavenumber was estimated by fitting a fourier transform for each latitude circle, level and monthly record.

We analyzed the data using the statistical programming language R [@R-base], using data.table and metR packages [@R-data.table; @R-metR] to read and transform it and ggplot2 package [@R-ggplot2] to make the plots. The source code is available as Figshare repository [@Campitelli2019-figshare].

# Results

```{r compute-qs-zw-rao}
qs <- rao_data %>% 
   .[, .(gh = mean(gh)), by = .(lon, lat, lev, month(time))] %>% 
   .[, FitWave(gh, 1:3), keyby = .(lat, lev, month)] %>% 
   addkey(k) %>% 
   .[, phase := phase*k] %>% 
   .[, c("r2") := NULL]

zw <- rao_data %>%
   .[, FitWave(gh, 1:3), by = .(lat, lev, time)] %>% 
   .[, .(amplitude = mean(amplitude),
         phase = as.numeric(mean(circular(phase*k)))), 
     keyby = .(k, lat, lev, month(time))] 
rao_qs <- rbindlist(list(QS = qs,
                         ZW = zw), 
                    use.names = TRUE,
                    idcol = "variable") 
rao_qs[, variable := factor(variable)]

math_levels <- c(TeX("Amplitude of quasi-stationary wave ($A_{QS}$)"),
                 TeX("Mean amplitude of zonal waves ($\\bar{A_{ZW}}$)"))
```

```{r}
computed <- "computed as the amplitude of the mean wave ($A_\\mathrm{QSk}$) and as the mean amplitude of the monthly waves ($\\overline{A_\\mathrm{ZW} }$)"
k_prefix <- AddPreffix("Wavenumber ")
```

```{r rao, fig.cap = glue("Seasonal cycle of amplitude of the geopotential planetary waves 1 to 3 {computed}."), fig.subcap=c("At 50\\degree N", "At 50\\degree S"), fig.width=5, fig.ncol=1}
plot_crossection <- function(k, lat, binwidth) {
   ks <- k
   lats <- lat
   gdata <- periodic(rao_qs[k == ks & lat == lats], month = 1:13)
   levels(gdata$variable) <- math_levels
   g <- periodic(gdata[k == ks], month = 1:13) %>% 
      ggplot(aes(month, lev)) +
      # geom_contour_fill(aes(z = value), binwidth = binwidth) +
      geom_contour2(aes(z = amplitude), binwidth = binwidth) +
      geom_text_contour(aes(z = amplitude), binwidth = binwidth, rotate = F, skip = 1,
                        stroke = 0.2, size = 2) +
      scale_y_level("Pressure (hPa)") +
      scale_x_continuous("Month", breaks = 1:13, 
                         labels = c(month.abb, "13" = "Jan"),
                         expand = c(0, 0), minor_breaks = NULL) +
      # scale_fill_viridis_c(guide = guide_colorstrip_bottom()) +
      # coord_fixed(1.7) +
      theme(aspect.ratio = 0.5) +
      facet_grid(k ~ variable,
                 labeller = labeller(k = k_prefix,
                                     variable = label_parsed))
   
   if (k != 3) {
      g <- g +
         theme(axis.text.x = element_blank(), axis.title.x = element_blank())
   }
   if (k != 1) {
      g <- g + 
         theme(axis.title.y = element_blank(),
               strip.text.x = element_blank(),
               plot.margin = unit(c(-0, 1, 1, 1), "lines"))
   }
   return(g)
}

binwidths <- c(50, 10, 5)
binwidths2 <- c(25, 10, 5)*2

lapply(1:3, function(x) plot_crossection(x, lats, binwidths2[x]))  %>% 
   Reduce("+", .) + 
   plot_layout(ncol = 1) 

lapply(1:3, function(x) plot_crossection(x, -lats, binwidths[x]))  %>% 
   Reduce("+", .) + 
   plot_layout(ncol = 1) 
```

Figure \ref{fig:rao} shows the seasonal cycle of the amplitude of planetary waves at `r lats`\degree S and `r lats`\degree N using monthly fields from the NCEP/NCAR reanalysis [@Kalnay1996] between `r period_rao[1]` and `r period_rao[2]`. We computed the left column ($A_\mathrm{QS}$) as the amplitude of the average geopotential field for each month, level and wavenumber, and the right column ($\overline{A_\mathrm{ZW}}$) as the average amplitude of the `r N` individual fields. 

Figure \ref{fig:rao1} shows that at `r lats`\degree N for the three wavenumbers $A_\mathrm{QS}$ and $\overline{A_\mathrm{ZW}}$ have a similar seasonal cycle with similar vertical extent. In the southern hemisphere, however, this is true only for wavenumber 1 (Figure \ref{fig:rao2}). $A_\mathrm{QS2}$ is much smaller than  $\overline{A_\mathrm{ZW2}}$ and its seasonal cycle is less defined. $A_\mathrm{QS3}$ has a smaller magnitude than $\overline{A_\mathrm{ZW3}}$ end even though their overall structure is similar (one relative maximum in February-March in the middle troposphere and another in July-August that extends to the lower stratosphere), they differ in the details. $A_\mathrm{QS3}$ has a local minimum in November that is absent in $\overline{A_\mathrm{ZW3}}$. The relative contribution of each wavenumber is also different. While $\overline{A_\mathrm{ZW2}}$ dominates over $\overline{A_\mathrm{ZW3}}$ in the stratosphere and is of similar magnitude in the troposphere, $A_\mathrm{QS3}$ dominates over $A_\mathrm{QS2}$ throughout the year and in every level except in the aforementioned November minimum. 


```{r stationarity, fig.cap = "Seasonal cycle of stationarity at 50\\degree S and 50\\degree N computed using Equation \\ref{eq:S}.", fig.align="center", fig.width=5}
black_contour <- 0.4
rao_qs %>% 
   dcast(lev + month + lat + k ~ variable, value.var = "amplitude") %>% 
   .[, S := QS/ZW] %>% 
   .[, lat_f := factor(lat, levels = c(lats, -lats))] %>% 
   ggplot(aes(month, lev)) +
   geom_contour_fill(aes(z = S), binwidth = 0.1) +
   geom_contour(aes(z = S), breaks = black_contour, color = "black", size = 0.2) +
   scale_y_level("Pressure (hPa)") +
   scale_x_continuous("Month", breaks = 1:13, 
                      labels = c(month.abb, "13" = "Jan"),
                      expand = c(0, 0), minor_breaks = NULL) +
   scale_fill_viridis_c(guide = guide_colorstrip_bottom(15),
                        breaks = MakeBreaks(0.1)) +
   theme(aspect.ratio = 0.5) +
   facet_grid(k ~ lat_f, labeller = labeller(k = k_prefix, 
                                             lat_f = LatLabel))
``` 

These observations are quantified in Figure \ref{fig:stationarity}, which shows $\hat{S}$ for wavenumbers 1 to 3 computed using Equation \ref{eq:S} at `r lats`\degree N and `r lats`\degree S. We separate between high and low stationarity with the ad-hoc threshold of `r black_contour` (black line in Figure \ref{fig:stationarity}).  

At `r lats`\degree N planetary waves 1, 2 and 3 are highly stationary in almost every month and level, and even more so planetary wave 1 at `r lats`\degree S. 

In the southern hemisphere, planetary wave 2 stationarity has a semiannual cycle. It reaches its maximum in April and in August-September, plummeting to a deep minimum in June. Planetary wave 3 stationarity peaks in February and slowly decreases towards a November deep minimum after witch increases sharply. 

```{r s-timeseries, fig.cap = "Quasi-stationary wave stationarity for wavenumbers 1 to 3.", fig.width=5}
pw <- rao_data %>%
   .[lev == 500, FitWave(gh, 1:3), by = .(lat, lev, time)]

pw[, phase.mean := mean.phase(amplitude, phase, k), 
   by = .(k, lat, lev, month(time))]
pw[, cor := cos(k*(phase - phase.mean))]
pw[, c("S", "S_se", "S_df") := predict(loess(cor ~ as.numeric(time), span = 10*12/.N,
                        weights = amplitude,
                        degree = 0), se = TRUE)[c("fit", "se.fit", "df")], 
   by = .(k, lat, lev)]
pw[, S_se1 := S_se*stats::qt(0.95 / 2 + .5, S_df)]

pw %>% 
   ggplot(aes(time, S)) +
   geom_ribbon(aes(ymin = S - S_se1,
                   ymax = S + S_se1, group = factor(k)),
               alpha = 0.2) +
   geom_line(aes(linetype = factor(k)), size = 1) +
   stat_subset(aes(subset = time == max(time),
                   # x = time + months(18),
                   label = k), geom = ggrepel::GeomTextRepel, 
               direction = "y", nudge_x = as.numeric(months(20))) +
   scale_y_continuous(TeX("\\hat{S}")) +
   scale_x_datetime("Time", breaks = function(x) pretty(x, 15), 
                    date_labels = "%y",
                    minor_breaks = NULL, 
                    expand = c(0, 0, 0, as.numeric(months(36)))) +
   scale_linetype_manual(guide = "none", values = c("1" = 1, "2" = 2, "3" = 3)) +
   coord_cartesian(ylim = c(0, 1)) +
   facet_wrap(~lat, ncol = 1, labeller = labeller(lat = LatLabel),
              as.table = FALSE)
```

Because we computed $\hat{S}$ using the whole period, it represents the mean stationarity between `r period_rao[1]` and `r period_rao[2]`. Figure \ref{fig:s-timeseries} shows how planetary wave stationarity changed for each wavenumber at each latitude. Planetary wave stationarity remained high and constant for wavenumbers 1 to 3 at `r lats`\degree N and 1 at `r lats`\degree S but wavenumbers 2 and 3 at `r lats`\degree S show interdecadal variations.

Planetary wave 2 stationarity oscillated around `r round(pw[lat == -lats & k == 2, mean(S)], 1)` with maximums in the 50's, 70's and 00's. Planetary wave 3 stationarity jumped from zero to more than 0.5 in less than five years in the 50's and then oscillated around `r round(pw[lat == -lats & k == 3 & year(time) > 1960, mean(S)], 1)` with a strong maximum in the late 80's. These could indicate inhomogeneities caused by changes in the observational network --routine satellite observations began in 1979-- but the absence of similar breaks for wavenumbers 1 or 2 suggest they represent real changes in the atmospheric circulation with unknown cause.

# Conclusions

We used a quantitative measure of planetary wave stationarity to show that in the southern hemisphere planetary wave 1 stationarity is high and constant throughout the year and period, while waves 2 and 3 vary both in intraseasonal and interdecadal timescales. Planetary wave 3 stationarity, in particular, increased dramatically in the 50's. In the northern hemisphere, planetary wave stationarity is higher and varies much less. 

Planetary waves can be both forced by the surface and excited by internal variability. Assuming that the later process will not result in a phase preference, higher stationarity would be evidence of stronger forcing or, more strictly, stronger forcing response. In the northern hemisphere, topography and thermal contrasts are the main forcings of planetary waves (REF), which explains their highly and not variable stationary nature. In the southern hemisphere, only planetary wave 1 seems to be the result of mainly surface forcings. Planetary waves 2 and 3 seem to be composed of a comparable mix of internal variability and surface forcing components. 

Interannual and intradecadal variability in planetary wave stationarity may serve to study variability in surface forcing responses caused either by the strength of the forcing or the sensitivity of mean state of the atmosphere.  

Since in the southern hemisphere the amplitude of the mean planetary wave can differ greatly from the mean amplitude of planetary waves, care must be taken when interpreting the literature. Some studies analyze the former [e.g. @Loon1972, @Quintanar1995a, @Raphael2004] while others analyze the later [e.g. @Rao2004, @Turner2017, @Irving2015]. For instance, @Irving2015 compare their planetary wave activity index with @Raphael2004's wave 3 index and conclude that the later cannot account for events with waves far removed from their climatological position. However, being an index of the zonal wave component in phase with the quasi-stationary wave, this is by design. 

The explorations of both zonal waves and quasi-stationary waves can lead to novel levels of analysis. For example, @Smith2012 used their phase relationship to show that linear interference between the zonal waves 1 and quasi-stationary wave 1 was related to vertical wave activity transport at the tropopause. Here, we showed it can be used to define a metric of stationarity of quasi-stationary waves, but other applications are also possible.

